The role of cognateness in non-native spoken word recognition

### 2024-05-08 manuscript/manuscript.tex
New analyses for Experiment 3

Data collection took place from June 04th, 2020 to June 25th, 2020. We
collected data from 71 British English native participants living in
United Kingdom (\emph{Mean} = 21.76 years, \emph{SD} = 2.15, Range =
18-26, 46, 25 female).

@@

Participants were recruited via Prolific (5£ compensation) and SONA
(compensation in academic credits), and gave informed consent before
providing any data and the study was conducted in accordance with
ethical standards of the Declaration of Helsinki and the protocol was by
the University of Oxford Medical Sciences Inter-Divisional Research
Ethics Committee (IDREC) (R60939/RE007). Participants were asked to
complete the experiment using a laptop in a quiet place with good
internet connection. We excluded data from participants that (a)
self-rated their oral and/or written skills in Catalan and Spanish, or
any other second language as four or higher in a five-point scale
(\emph{n} = 2), (b) were diagnosed with a language (\emph{n} = 1), or
(c) did not contribute more than 80\% of valid trials (\emph{n} = 9).

@@

Responses given by English participants to Catalan presented words were
5.35 characters long on average (\emph{Median} = 5, \emph{SD} = 1.79,
Range = 1-14), while their translations to Spanish responses were 5.57
characters long on average (\emph{Median} = 5, \emph{SD} = 1.97, Range =
2-21).

@@

\toprule
\multicolumn{1}{l}{} & \multicolumn{2}{c}{Correct responses} & \multicolumn{2}{c}{Incorrect responses} \\ 
\cmidrule(lr){2-3} \cmidrule(lr){4-5}
\multicolumn{1}{l}{} & Correct & Typo & Wrong translation & False friend \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{Experiment 1} \\ 
\midrule\addlinespace[2.5pt]
cat-ENG & $429$ ($16.47\%$) & $11$ ($0.42\%$) & $2,082$ ($79.95\%$) & $82$ ($3.15\%$) \\ 
spa-ENG & $374$ ($14.37\%$) & $2$ ($0.08\%$) & $2,117$ ($81.36\%$) & $109$ ($4.19\%$) \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{Experiment 2} \\ 
\midrule\addlinespace[2.5pt]
cat-ENG & $505$ ($18.41\%$) & $7$ ($0.26\%$) & $2,052$ ($74.81\%$) & $179$ ($6.53\%$) \\ 
spa-ENG & $630$ ($20.03\%$) & $6$ ($0.19\%$) & $2,352$ ($74.79\%$) & $157$ ($4.99\%$) \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{Experiment 3} \\ 
\midrule\addlinespace[2.5pt]
cat-SPA & $780$ ($46.93\%$) & $20$ ($1.20\%$) & $736$ ($44.28\%$) & $126$ ($7.58\%$) \\ 
\bottomrule

@@

We modelled the probability of participants guessing the correct
translation of each presented word using a generalised multilevel
Bayesian regression model with a Bernoulli logit link distribution. We
included as fixed effects the intercept, the main effects of
\emph{Frequency}, \emph{Cognateness}, and \emph{CLPN}, and the two-way
interaction between \emph{Cognateness} and \emph{CLPN}. We also included
participant-level random intercepts and slopes for the the main effects
and the interaction. \textbf{?@eq-model} shows a formal description of
the model.

To test the practical relevance of each predictor we followed
(\citeproc{ref-krushke2018bayesian}{\textbf{krushke2018bayesian?}}). We
first specified a region of practical equivalence (\emph{ROPE}) around
zero. This area indicates the values of the regression coefficients that
we considered as equivalent to zero. We then computed the 95\% posterior
credible intervals (CrI) of the regression coefficients of interest.
Finally, we calculated the proportion of the 95\% CrI that fell inside
the ROPE. This proportion indicates the probability that the true value
of the coefficient is equivalent to zero.

\textbf{?@tbl-results-1} shows a summary of participants' accuracy
across Experiments 1, 2, and 3. Overall, participants responded less
accurately to words with more CLPNs than to words with fewer CLPNs,
regardless of the amount of phonological similarity between the
presented word and its translation. This is indicated by the size of the
regression coefficient of the two-way interaction between
\emph{Cognateness} and \emph{CLPN} (\(\beta\) = -0.629, 95\% CrI =
{[}-0.982, -0.344{]}, \emph{p}(ROPE) = 0). As anticipated, participants'
performance benefited from an increase in \emph{Cognateness} (\(\beta\)
= 7.11, 95\% CrI = {[}6.567, 7.635{]}, \emph{p}(ROPE) = 0), while the
number of \emph{CLND} had the opposite effect (\(\beta\) = 0.007, 95\%
CrI = {[}-0.107, 0.103{]}, \emph{p}(ROPE) = 0.927). A model including
\emph{Group} as a main effect did not improve the predictive performance
of the model (\(\Delta_{ELPD}\) = -1.33, \(SE\) = 1.14), indicating that
participants in both groups performed equivalently.
Figure~\ref{fig-epreds-1} illustrates the posterior of the average
predictions of the model for words with different values of
\emph{Cognateness} and \emph{CLPN}.

@@

\toprule
\multicolumn{1}{l}{} &  & \multicolumn{3}{c}{Valid trials} & \multicolumn{3}{c}{Accuracy (\%)} \\ 
\cmidrule(lr){3-5} \cmidrule(lr){6-8}
\multicolumn{1}{l}{} & N & Mean ± SD & N trials & Range & Mean ± SD & SE & Range \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{8}{l}{Experiment 1} \\ 
\midrule\addlinespace[2.5pt]
spa-ENG & $27$ & $96.37$ ± $2.88$ ± $2.88$ & $2,602$ & $87$–$98$ & $15.86$ ± $5.20$ & $3.05$ & $8.82$–$28.71$ \\ 
cat-ENG & $32$ & $81.38$ ± $3.17$ ± $3.17$ & $2,604$ & $71$–$83$ & $18.48$ ± $4.89$ & $3.27$ & $10.84$–$32.56$ \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{8}{l}{Experiment 2} \\ 
\midrule\addlinespace[2.5pt]
spa-ENG & $31$ & $101.45$ ± $1.80$ ± $1.80$ & $3,145$ & $92$–$102$ & $21.35$ ± $8.10$ & $3.83$ & $6.60$–$44.34$ \\ 
cat-ENG & $32$ & $85.72$ ± $0.52$ ± $0.52$ & $2,743$ & $84$–$86$ & $20.06$ ± $5.37$ & $3.55$ & $10.00$–$30.00$ \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{8}{l}{Experiment 3} \\ 
\midrule\addlinespace[2.5pt]
cat-SPA & $21$ & $79.14$ ± $3.14$ ± $3.14$ & $1,662$ & $72$–$82$ & $48.30$ ± $5.29$ & $10.54$ & $38.27$–$58.97$ \\ 
\bottomrule

@@

@@

@@

\textbf{?@tbl-results-1} shows a summary of participants' accuracy
across Experiments 1, 2, and 3. Overall, participants responded less
accurately to words with more CLPNs than to words with fewer CLPNs,
regardless of the amount of phonological similarity between the
presented word and its translation. This is indicated by the size of the
regression coefficient of the two-way interaction between
\emph{Cognateness} and \emph{CLPN} (\(\beta\) = -0.584, 95\% CrI =
{[}-0.852, -0.331{]}, \emph{p}(ROPE) = 0). As anticipated, participants'
performance benefited from an increase in \emph{Cognateness} (\(\beta\)
= 6.926, 95\% CrI = {[}6.4, 7.461{]}, \emph{p}(ROPE) = 0), while the
number of \emph{CLND} had the opposite effect (\(\beta\) = 0, 95\% CrI =
{[}-0.113, 0.085{]}, \emph{p}(ROPE) = 0.95). A model including
\emph{Group} as a main effect did not improve the predictive performance
of the model (\(\Delta_{ELPD}\) = -6.56, \(SE\) = 3.21), indicating that
participants in both groups performed equivalently.
Figure~\ref{fig-epreds-1} illustrates the posterior of the average
predictions of the model for words with different values of
\emph{Cognateness} and \emph{CLPN}.

@@

@@

@@

@@

@@

\textbf{?@tbl-results-1} shows a summary of participants' accuracy
across Experiments 1, 2, and 3. Overall, participants responded less
accurately to words with more CLPNs than to words with fewer CLPNs,
regardless of the amount of phonological similarity between the
presented word and its translation. This is indicated by the size of the
regression coefficient of the two-way interaction between
\emph{Cognateness} and \emph{CLPN} (\(\beta\) = -0.409, 95\% CrI =
{[}-0.895, 0.111{]}, \emph{p}(ROPE) = 0.081). As anticipated,
participants' performance benefited from an increase in
\emph{Cognateness} (\(\beta\) = 9.274, 95\% CrI = {[}8.473, 10.269{]},
\emph{p}(ROPE) = 0), while the number of \emph{CLND} had the opposite
effect (\(\beta\) = -0.069, 95\% CrI = {[}-0.353, 0.192{]},
\emph{p}(ROPE) = 0.492). Figure~\ref{fig-epreds-1} illustrates the
posterior of the average predictions of the model for words with
different values of \emph{Cognateness} and \emph{CLPN}.

@@

@@

>>>>

Participants were recruited via Prolific (5£ compensation) and SONA
(compensation in academic credits), and gave informed consent before
providing any data and the study was conducted in accordance with
ethical standards of the Declaration of Helsinki and the protocol was by
the University of Oxford Medical Sciences Inter-Divisional Research
Ethics Committee (IDREC) (R60939/RE007). Participants were asked to
complete the experiment using a laptop in a quiet place with good
internet connection. Data collection took place from June 04th, 2020 to
June 25th, 2020. We collected data from 71 British English native
participants living in United Kingdom (\emph{Mean} = 21.76 years,
\emph{SD} = 2.15, Range = 18-26, 46, 25 female).

@@

@@

We modelled the probability of participants guessing the correct
translation of each presented word using a generalised multilevel
Bayesian regression model with a Bernoulli logit link distribution. We
included as fixed effects the intercept, the main effects of
\emph{Frequency}, \emph{Cognateness}, and \emph{CLPN}, and the two-way
interaction between \emph{Cognateness} and \emph{CLPN}. We also included
participant-level random intercepts and slopes for the the main effects
and the interaction. Eq. 1 shows a formal description of the model.

{\centering \includegraphics{_assets/img/model.PNG}

\caption{Eq. 1}

To test the practical relevance of each predictor we followed
(\citeproc{ref-kruschke2018bayesian}{Kruschke \& Liddell, 2018}). We
first specified a region of practical equivalence (\emph{ROPE}) around
zero ({[}-0.1, +0.1{]}, in the logit scale). This area indicates the
values of the regression coefficients that we considered as equivalent
to zero. We then computed the 95\% posterior credible intervals (CrI) of
the regression coefficients of interest. Finally, we calculated the
proportion of the 95\% CrI that fell inside the ROPE. This proportion
indicates the probability that the true value of the coefficient is
equivalent to zero.

We collected data for a total of 6,446 trials completed by 72 unique
participants. Of those trials, 2,988 were provided by 36 unique
participants who listened to Catalan words, and 3,458 trials were
provided by 36 unique participants who listened to Spanish words. We
excluded trials in which participants did not enter any text (\emph{n} =
72), in which a response in a language other than English was provided
(e.g., \texttt{agua}, \emph{n} = 51), in which participants did not
provide a whole word (e.g., \texttt{f}, \emph{n} = 5), and in which
participants added comments to the experimenter (e.g., \texttt{unsure},
(\emph{n} = 13). In addition, we excluded data from participants that
self-rated their oral and/or written skills in Catalan and Spanish, or
any other second language as four or higher in a five-point scale
(\emph{n} = 2), were diagnosed with a language (\emph{n} = 1), or did
not contribute more than 80\% of valid trials (\emph{n} = 9). After
applying trial-level and participant-level inclusion criteria, the
resulting dataset included 5,204 trials provided by 54 unique
participants. Of those trials, 2,602 were provided by 27 unique
participants who listened to Catalan words, and 2,604 trials were
provided by 32 unique participants who listened to Spanish words.
Responses given by English participants to Catalan presented words were
5.35 characters long on average (\emph{Median} = 5, \emph{SD} = 1.79,
Range = 1-14), while their translations to Spanish responses were 5.57
characters long on average (\emph{Median} = 5, \emph{SD} = 1.97, Range =
2-21).

@@

\toprule
\multicolumn{1}{l}{} & \multicolumn{2}{c}{Correct responses} & \multicolumn{2}{c}{Incorrect responses} \\ 
\cmidrule(lr){2-3} \cmidrule(lr){4-5}
\multicolumn{1}{l}{} & Correct & Typo & Wrong translation & False friend \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{Experiment 1} \\ 
\midrule\addlinespace[2.5pt]
cat-ENG & $429$ ($16.47\%$) & $11$ ($0.42\%$) & $2,082$ ($79.95\%$) & $82$ ($3.15\%$) \\ 
spa-ENG & $374$ ($14.37\%$) & $2$ ($0.08\%$) & $2,117$ ($81.36\%$) & $109$ ($4.19\%$) \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{Experiment 2} \\ 
\midrule\addlinespace[2.5pt]
cat-SPA & $780$ ($46.93\%$) & $20$ ($1.20\%$) & $736$ ($44.28\%$) & $126$ ($7.58\%$) \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{5}{l}{Experiment 3} \\ 
\midrule\addlinespace[2.5pt]
cat-ENG & $477$ ($18.01\%$) & $7$ ($0.26\%$) & $1,986$ ($75.00\%$) & $178$ ($6.72\%$) \\ 
spa-ENG & $590$ ($19.53\%$) & $6$ ($0.20\%$) & $2,294$ ($75.94\%$) & $131$ ($4.34\%$) \\ 
\bottomrule

@@

Table~\ref{tbl-results} shows a summary of participants' accuracy across
Experiments 1, 2, and 3. MCMC chains in the model showed strong evidence
of convergence (\(\hat{R}<1.01\)) (see Appendix 2 for more detailed
model diagnostics). Participants translating Catalan words and
participants translating Spanish words performed equivalently, as
indicated by the regression coefficient of \emph{Group} (\(\beta\) =
-0.199, 95\% CrI = {[}-0.512, 0.129{]}, \emph{p}(ROPE) = 0.238).
Overall, participants responded less accurately to words with more CLPNs
than to words with fewer CLPNs, regardless of the amount of phonological
similarity between the presented word and its translation. This is
indicated by the size of the regression coefficient of the two-way
interaction between \emph{Cognateness} and \emph{CLPN} (\(\beta\) =
-0.653, 95\% CrI = {[}-0.973, -0.313{]}, \emph{p}(ROPE) = 0). As
anticipated, participants' performance benefited from an increase in
\emph{Cognateness} (\(\beta\) = 7.133, 95\% CrI = {[}6.606, 7.735{]},
\emph{p}(ROPE) = 0), while the number of \emph{CLND} had the opposite
effect (\(\beta\) = 0.009, 95\% CrI = {[}-0.11, 0.111{]}, \emph{p}(ROPE)
= 0.925). Figure~\ref{fig-epreds-1} illustrates the posterior of the
average predictions of the model for words with different values of
\emph{Cognateness} and \emph{CLPN}. Figure~\ref{fig-coefs} shows a
graphic summary of the posterior distribution of the regression
coefficients of interest.

@@

\toprule
\multicolumn{1}{l}{} &  & \multicolumn{3}{c}{Accuracy (\%)} & \multicolumn{3}{c}{Valid trials} \\ 
\cmidrule(lr){3-5} \cmidrule(lr){6-8}
\multicolumn{1}{l}{} & N & Mean ± SD & SE & Range & Mean ± SD & N trials & Range \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{8}{l}{Experiment 1} \\ 
\midrule\addlinespace[2.5pt]
spa-ENG & $27$ & $15.86$ ± $5.20$ & $3.05$ & $8.82$–$28.71$ & $96.37$ ± $2.88$ ± $2.88$ & $2,602$ & $87$–$98$ \\ 
cat-ENG & $32$ & $18.48$ ± $4.89$ & $3.27$ & $10.84$–$32.56$ & $81.38$ ± $3.17$ ± $3.17$ & $2,604$ & $71$–$83$ \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{8}{l}{Experiment 2} \\ 
\midrule\addlinespace[2.5pt]
cat-SPA & $21$ & $48.30$ ± $5.29$ & $10.54$ & $38.27$–$58.97$ & $79.14$ ± $3.14$ ± $3.14$ & $1,662$ & $72$–$82$ \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{8}{l}{Experiment 3} \\ 
\midrule\addlinespace[2.5pt]
spa-ENG & $31$ & $20.92$ ± $8.29$ & $3.76$ & $5.88$–$44.12$ & $97.45$ ± $1.80$ ± $1.80$ & $3,021$ & $88$–$98$ \\ 
cat-ENG & $32$ & $19.74$ ± $4.94$ & $3.49$ & $10.34$–$27.91$ & $82.75$ ± $0.44$ ± $0.44$ & $2,648$ & $82$–$83$ \\ 
\bottomrule

@@

In the present experiment, we investigated the extent to which
cognateness (phonological similarity between translation equivalents) is
sufficient for successful word translation, in the absence of semantic
knowledge. We tested two groups of monolingual British English natives
in a translation task that involved words in Catalan or Spanish, two
languages participants reported having no familiarity with. Participants
benefited strongly from cognateness, when the correct translation of the
presented words in Catalan or Spanish had few English phonological
neighbours with higher lexical frequency. This suggests that, in the
absence of distractors, even naïve participants can efficiently use
phonological similarity to succeed in a translation task. Our results
suggest that word-to-word connections at the phonological level might
play a strong role during L2 speech comprehension, specially in
low-proficiency listeners.

@@

\subsubsection{Participants}\label{participants-1}

Participants in Spain were contacted via announcements in Faculties, and
were compensated 5€ or an Amazon voucher for the same value.
Participants gave informed consent before providing any data and the
study was conducted in accordance with ethical standards of the
Declaration of Helsinki and the protocol was approved by the Drug
Research Ethical Committee (CEIm) of the IMIM Parc de Salut Mar
(2020/9080/I). Data collection took place from June 08th, 2020 to June
28th, 2020. We collected data from 33 Spanish native participants living
in Spain (\emph{Mean} = 21.85 years, \emph{SD} = 3, Range = 18-33, 28, 5
female).

\subsubsection{Stimuli}\label{stimuli-1}

Same list of Catalan stimuli as in Experiment 1.

\subsubsection{Procedure}\label{procedure-1}

\subsubsection{Data analysis}\label{data-analysis-1}

Same approach as in Experiment 1.

@@

We collected data for a total of 5,412 trials completed by 33 unique
participants. We excluded trials in which participants did not enter any
text (\emph{n} = 44), in which a response in a language other than
English was provided (e.g., \texttt{agua}, \emph{n} = 51), in which
participants did not provide a whole word (e.g., \texttt{f}, \emph{n} =
7), and in which participants added comments to the experimenter (e.g.,
\texttt{unsure}, (\emph{n} = 1). In addition, we excluded data from
participants that self-rated their oral and/or written skills in Catalan
and Spanish, or any other second language as four or higher in a
five-point scale (\emph{n} = 22), were diagnosed with a language
(\emph{n} = 1), or did not contribute more than 80\% of valid trials
(\emph{n} = 9). After applying trial-level and participant-level
inclusion criteria, the resulting dataset included 1,662 trials provided
by 42 unique participants. Of those trials, 1,662 were provided by 21
unique participants who listened to Catalan words, and 1,662 trials were
provided by 21 unique participants who listened to Spanish words.
Responses given by participants were 5.6 characters long on average
(\emph{Median} = 5, \emph{SD} = 1.6, Range = 2-12).

\textbf{?@tbl-results-1} shows a summary of participants' accuracy
across Experiments 1, 2, and 3. MCMC chains in the model showed strong
evidence of convergence (\(\hat{R}<1.01\)) (see Appendix 2 for more
detailed model diagnostics). Overall, participants responded less
accurately to words with more CLPNs than to words with fewer CLPNs,
regardless of the amount of phonological similarity between the
presented word and its translation. This is indicated by the size of the
regression coefficient of the two-way interaction between
\emph{Cognateness} and \emph{CLPN} (\(\beta\) = -0.409, 95\% CrI =
{[}-0.895, 0.111{]}, \emph{p}(ROPE) = 0.081). As anticipated,
participants' performance benefited from an increase in
\emph{Cognateness} (\(\beta\) = 9.274, 95\% CrI = {[}8.473, 10.269{]},
\emph{p}(ROPE) = 0), while the number of \emph{CLND} had the opposite
effect (\(\beta\) = -0.069, 95\% CrI = {[}-0.353, 0.192{]},
\emph{p}(ROPE) = 0.492). Figure~\ref{fig-epreds-1} illustrates the
posterior of the average predictions of the model for words with
different values of \emph{Cognateness} and \emph{CLPN}.

@@

Experiment 2 was an extension of Experiment 1 to a population of
monolinguals whose native language is typologically similar to the
presented language. Particularly, we presented Catalan words to Spanish
native speakers who were reportedly unfamiliar with Catalan. Our results
indicate a similar pattern of results as those in Experiment 1:
participants were able to provide correct translations of presented
Catalan words, provided that the Catalan words shared some degree of
phonological similarity with their Spanish translation, and that the
number of phonological neighbours with higher lexical frequency was
reduced. In contrasts with the results in Experiment 1, the positive
impact of phonological similarity on participants' performance in
Experiment 2 was more resilient to the interference of phonological
neighbours. English natives in Experiment 1 provided significantly less
accuracy responses when more than four phonological neighbors were
present (even when translating high-similarity words), compared to when
only one or neighbour were present. Spanish participants in Experiment 2
benefited from phonological similarity, even when eight neighbours were
present. Spanish participants' performance declined after 8 neighbours,
and was evident at 12 neighbours. Overall, this suggests that
participants in Experiment 2, who were natives of a typologically
similar language (Spanish) to the presented language (Catalan) benefited
more strongly from phonological similarity than participants in
Experiment 1, who were natives of typologically less similar language
(English) to the presented language (Catalan, Spanish).

Participants from both Experiment 1 and 2 benefited strongly from
phonological similarity to correctly translate words from a non-native,
reportedly unfamiliar language. This pattern of results holds for most
of the presented stimuli, but some low-similarity Catalan and Spanish
words were responded to surprisingly accurately by English listeners.
Given that participants were reportedly unfamiliar with both languages,
it was expected that participants would be very unlikely to provide
correct translations for words sharing little to no phonological
similarity to their correct translation. Table~\ref{tbl-surprises} a
list of Catalan and Spanish words to which participants provided
responses with \(\geq\) 10 average accuracy.

\begin{longtable}{l|rr}

\caption{\label{tbl-surprises}List of items with unexpectedly high
accuracy: the Levenshtein similarityscore betwen th presented word (in
Catalan or Spanish) and their correct Enlgish or Spanish translation is
zero, but participants, who are reportedly unfamiliar with the presented
language, were on average \textgreater10\% likely to guess the correct
translation.}

\toprule
\multicolumn{1}{l}{} & Accuracy (\%) & SE \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{3}{l}{Experiment 1 - cat-ENG} \\ 
\midrule\addlinespace[2.5pt]
cavall /kəβaʎ/ - horse /hɔs/ & $17.14$ & $6.37$ \\ 
llibre /ʎiβɾə/ - book /bʊk/ & $17.14$ & $6.37$ \\ 
camisa /kəmizə/ - shirt /ʃɜt/ & $16.67$ & $6.21$ \\ 
poma /poma/ - apple /æpl/ & $16.67$ & $6.21$ \\ 
cama /kamə/ - leg /lɛg/ & $11.11$ & $5.24$ \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{3}{l}{Experiment 1 - spa-ENG} \\ 
\midrule\addlinespace[2.5pt]
pantalón /paŋtalon/ - trousers /traʊzəz/ & $77.42$ & $7.51$ \\ 
naranja /naɾaŋxa/ - orange /ɒrɪnʤ/ & $41.94$ & $8.86$ \\ 
leche /leʧe/ - milk /mɪlk/ & $35.48$ & $8.59$ \\ 
toro /toɾo/ - bull /bʊl/ & $33.33$ & $8.61$ \\ 
libro /liβɾo/ - book /bʊk/ & $30.00$ & $8.37$ \\ 
cebra /θebɾa/ - zebra /zibrə/ & $29.03$ & $8.15$ \\ 
pan /pan/ - bread /brɛd/ & $29.03$ & $8.15$ \\ 
pollo /poʎo/ - chicken /ʧɪkɪn/ & $26.67$ & $8.07$ \\ 
jirafa /xiɾafa/ - giraffe /ʤɪrɑf/ & $20.69$ & $7.52$ \\ 
perro /pero/ - dog /dɒɡ/ & $16.13$ & $6.61$ \\ 
pluma /pluma/ - feather /fɛðə/ & $16.13$ & $6.61$ \\ 
puerta /pwerta/ - door /dɔ/ & $16.13$ & $6.61$ \\ 
pie /pje/ - foot /fʊt/ & $12.90$ & $6.02$ \\ 
caballo /kaβaʎo/ - horse /hɔs/ & $10.34$ & $5.66$ \\ 
bocadillo /bokadiʎo/ - sandwich /sænwɪʤ/ & $10.00$ & $5.48$ \\ 
globo /ɡloβo/ - balloon /bəlun/ & $10.00$ & $5.48$ \\ 
\midrule\addlinespace[2.5pt]
\multicolumn{3}{l}{Experiment 2 - cat-SPA} \\ 
\midrule\addlinespace[2.5pt]
fulla /fuʎə/ - hoja /oxa/ & $30.43$ & $9.59$ \\ 
ull /uʎ/ - ojo /oxo/ & $21.74$ & $8.60$ \\ 
got /gɔt/ - vaso /baso/ & $20.00$ & $8.00$ \\ 
entrepà /entɾəpa/ - bocadillo /bokadiʎo/ & $13.04$ & $7.02$ \\ 
mirall /miɾaʎ/ - espejo /espexo/ & $12.50$ & $6.75$ \\ 
\bottomrule

It is likely that participants had prior knowledge of these words
despite having reported little to no familiarity with the presented
language (Catalan or Spanish). One possibility is that participants had
previously encountered these words embedded in English linguistic input.
Spanish words percolate English speech with relative frequency, via
different sources such as popular culture, songs, TV programs, etc. In
addition, words from languages other than Spanish, but with high
similarity to the Spanish words (e.g., cognates from Italian or French)
might appear in English speech as well. Such prior knowledge might not
be specific to the low-similarity words highlighted before. Participants
may also have had prior knowledge about higher-similarity words, which
could have contributed to participants responding to such words more
accurately than without such prior knowledge. In the case of
higher-similarity words, it is more difficult to disentangle the extent
to which participants' accuracy is a function of pure phonological
similarity, or prior knowledge they had about the meaning of Spanish
words. To investigate this issue, we run Experiment 3.

{[}ADD HERE LEXICAL FREQUENCY OF CATALAN AND SPANISH WORDS IN ENGLISH,
AND OF CATALAN WORDS IN SPANISH{]}

@@

Experiment 3 is a replication of Experiment 1, in which we collected
additional data about participants' prior familiarity with the presented
Catalan and Spanish words, in addition to the same translation task
presented to participants in Experiment 1.

@@

\subsubsection{Participants}\label{participants-2}

Data collection took place from October 22th, 2022 to October 23th,
2022. We collected data from 64 British English native participants
living in United Kingdom (\emph{Mean} = 22.02 years, \emph{SD} = 2.49,
Range = 18-26, 36, 28 female).

Participants were recruited via Prolific (5£ compensation) and SONA
(compensation in academic credits), and gave informed consent before
providing any data and the study was conducted in accordance with
ethical standards of the Declaration of Helsinki and the protocol was by
the University of Oxford Medical Sciences Inter-Divisional Research
Ethics Committee (IDREC) (R60939/RE007). Participants were asked to
complete the experiment using a laptop in a quiet place with good
internet connection. We excluded data from participants that (a)
self-rated their oral and/or written skills in Catalan and Spanish, or
any other second language as four or higher in a five-point scale
(\emph{n} = 2), (b) were diagnosed with a language (\emph{n} = 0), or
(c) did not contribute more than 80\% of valid trials (\emph{n} = 1).

\subsubsection{Stimuli}\label{stimuli-2}

@@

\subsubsection{Procedure}\label{procedure-2}

{[}TO BE COMPLETED{]}

\subsubsection{Data analysis}\label{data-analysis-2}

Same approach as in Experiment 1.

@@

We collected data for a total of 6,016 trials completed by 64 unique
participants. Of those trials, 2,752 were provided by 32 unique
participants who listened to Catalan words, and 3,264 trials were
provided by 32 unique participants who listened to Spanish words. We
excluded trials in which participants did not enter any text (\emph{n} =
), in which a response in a language other than English was provided
(e.g., \texttt{agua}, \emph{n} = ), in which participants did not
provide a whole word (e.g., \texttt{f}, \emph{n} = ), and in which
participants added comments to the experimenter (e.g., \texttt{unsure},
(\emph{n} = ). In addition, we excluded data from participants that
self-rated their oral and/or written skills in Catalan and Spanish, or
any other second language as four or higher in a five-point scale
(\emph{n} = 2), were diagnosed with a language (\emph{n} = 0), or did
not contribute more than 80\% of valid trials (\emph{n} = 1). After
applying trial-level and participant-level inclusion criteria, the
resulting dataset included 6,290 trials provided by 62 unique
participants. Of those trials, 3,145 were provided by 31 unique
participants who listened to Catalan words, and 2,743 trials were
provided by 32 unique participants who listened to Spanish words.

Responses given by English participants to Catalan presented words were
5.52 characters long on average (\emph{Median} = 5, \emph{SD} = 1.75,
Range = 1-17), while their translations to Spanish responses were 5.41
characters long on average (\emph{Median} = 5, \emph{SD} = 1.77, Range =
1-20).

MCMC chains in the model showed strong evidence of convergence
(\(\hat{R}<1.01\)) (see Appendix 2 for more detailed model diagnostics).
Overall, participants reported prior knowledge more often for that
Spanish words with unexpectedly high accuracy (see Discussion in
Experiment 2) than for words with expected accuracy (see
Figure~\ref{fig-knowledge}). Participants reported prior knowledge of
Catalan words with unexpected accuracy as often as those with expected
accuracy. This suggests that participants in Experiment 1 may have
relied, to some extent, on their prior knowledge about form-meaning
mappings to correctly translate some Spanish words. To isolate such an
effect of prior Spanish knowledge, we run the same analysis as in
Experiment 1 on the newly collected translations from Experiment 3, now
excluding responses to words in which participants reported prior
knowledge.

\includegraphics{manuscript_files/figure-pdf/fig-knowledge-1.pdf}

\caption{\label{fig-knowledge}Catalan/Spanish prior word knowledge as
reported by English native participants in Experiment 3. (A) Average
proportion of participants that reported prior knowledge for words with
surprisingly high accuracy (no phonological similarity with the correct
translation, accuracy higher than 10\%), and for words with expected
accuracy (low similarity-low accuracy, or high similarity-high
accuracy). (B) Average accuracy for words with expected and unexpected
accuracy.}

Participants translating Catalan words and participants translating
Spanish words, as indicated by the regression coefficient of
\emph{Group} (\(\beta\) = -0.049, 95\% CrI = {[}-0.388, -0.32{]},
\emph{p}(ROPE) = 0.426). Overall, participants responded less accurately
to words with more CLPNs than to words with fewer CLPNs, regardless of
the amount of phonological similarity between the presented word and its
translation. This is indicated by the size of the regression coefficient
of the two-way interaction between \emph{Cognateness} and \emph{CLPN}
(\(\beta\) = -0.665, 95\% CrI = {[}-1.068, -0.32{]}, \emph{p}(ROPE) =
0.002). As anticipated, participants' performance benefited from an
increase in \emph{Cognateness} (\(\beta\) = 7.585, 95\% CrI = {[}7.043,
8.24{]}, \emph{p}(ROPE) = 0), while the number of \emph{CLND} had the
opposite effect (\(\beta\) = -0.049, 95\% CrI = {[}-0.203, 0.086{]},
\emph{p}(ROPE) = 0.753). Figure~\ref{fig-epreds-1} illustrates the
posterior of the average predictions of the model for words with
different values of \emph{Cognateness} and \emph{CLPN}.

@@

\section{Joint analyses}\label{joint-analyses}

Across Experiments 1 and 3, we found strong evidence that participants
efficiently exploited phonological similarity to provide accurate
translations for words in an unfamiliar language, provided that few
phonological neighbours of higher lexical frequency were present.
Figure~\ref{fig-coefs} summarizes the posterior distribution of the
regression coefficients of the models in Experiments 1 to 3.

\captionsetup{labelsep=none}\includegraphics{manuscript_files/figure-pdf/fig-coefs-1.pdf}

\caption{\label{fig-coefs}}

\section{General discussion}\label{general-discussion}

@@

\bibitem[\citeproctext]{ref-kruschke2018bayesian}
Kruschke, J. K., \& Liddell, T. M. (2018). The bayesian new statistics:
Hypothesis testing, estimation, meta-analysis, and power analysis from a
bayesian perspective. \emph{Psychonomic Bulletin \& Review}, \emph{25},
178--206.
